\chapter{Other topics}
\section{Virtualization}
The idea of virtualization is to put a layer between the implementation of something ad its interface. An application of virtualization are virtual file systems. The same idea can be applied to virtualize access to hardware: the software in charge of this is called virtual machine manager (VMM) or hypervisor. The OS that interfaces directly to the hardware is called host, while the ones running through a VMM are called guests. There are different types of VMMs:
\begin{itemize}
    \item Type 0 VMM: the VMM is implemented directly in hardware
    \item Type 1 VMM: the VMM is an OS-like program which allows execution of virtual machines and it runs at boot
    \item Type 2 VMM: the VMM is run as a normal user process on a real operating system. This solution is the one that gives the less freedom, since the host OS regulates the access to resources, the processor used, the amount of CPU time etc.
\end{itemize}

There are other variations of VMMs, namely:
\begin{itemize}
    \item Paravirtualization: the guest operating system is optimized to work well with the VMM
    \item Emulators: allow applications written for one hardware
    environment to run on a very different hardware environment
    \item Application containment: not actual virtualization, provides virtualization-like features by segregating applications from the operating system, making them more secure,
    manageable
\end{itemize}

\subsubsection{Running Mode}
The VMM is run in user mode on the host OS, but the guest OS need s to run some software in kernel mode. This means that the VMM needs to virtualize the kernel mode, which is known as trap-and-emulate.

\subsection{Containers}
Containers are a form of operating system virtualization. Inside a container are all the necessary executables, binary code, libraries, and configuration files to execute a software. The main difference with virtual machines is that they do not contain an OS. They still require a translation layer to interface with the host operating system, such as the Docker engine.

\subsection{Other applications of virtualization}
Another application of the virtualization idea is the programming environment virtualization. In this case the programming language is designed to run within a custom-built virtualized environment. For example Java has many features that depend on running in Java Virtual Machine (JVM).

\section{Security}
A system is said to be secure if resources are accesses and used as intended under all circumstances. \bold{Intruders} attempt to breach security. A \bold{threat} is a potential security violation. An \bold{attack} is an attempt to brach security, which can be accidental or malicious.

There are different types of security violations: breach of confidentiality (unauthorized access to data), breach of integrity (unauthorized modification/destruction of data), breach of availability (system/service is not ready for users).

There are different types of attacks: ransomware, replay attack, MITM attack, session hijacking, privilege escalation, trojan, backdoor, malware, spyware.

Security measures against attacks can be implemented at all layers of the system to be effective (physical, application, OS or network layer). Security is as weak as the weakest link in the chain.

Cryptography can be used as a tool to improve security (ex. authentication and communications). It can be implemented at various layers of the ISO reference model (network layer - IPSec, transport layer - SSL aka TLS).

Firewalls are software or hardware solutions that limit network access between two security domains.

\subsection{Principles of protection}
A very important principle of protection is the principle of least privilege. This principle states that programs, users and systems should be given just enough privileges to perform their tasks. Properly set permissions can limit damage if entity has a bug or gets abused. Privileges can be set statically or dynamically. Rough-grained privilege management is simpler, but less effective. Fine-grained management is more complex and adds more overhead, but is more protective. Additionally, an audit log can be maintained to record all sensitive activities.

\subsection{Protection methods}
Protection rings can be established: components are ordered by amount of privilege and are protected from each other. For example, the kernel is in one ring and user applications in another. This privilege separation requires hardware support.

Other protection methods are sandboxing (running processes in a limited environment) and code signing (verifying that the software is genuine by checking its signature).

\section{I/O systems}
\subsubsection{I/O communication}
The communication with I/O devices can be done in two ways:
\begin{itemize}
    \item Polling: the device sets a busy bit when it is communicating, the host periodically checks it waits until the device is not busy (host is busy waiting) and sets a command-ready bit when it wants to receive data.
    \item Interrupts: I/O device can trigger an interrupt line, which is connected directly to the CPU. The line is checked by the processor after every instruction and each interrupt is managed by the interrupt handler. The interrupt handler can ignore (maskable interrupts) or give priority to some interrupts. The interrupt vector to dispatch the interrupt to correct handler. The interrupt mechanism is also used for software exceptions and page faults (traps).
\end{itemize}

\subsubsection{Direct memory access}
Direct memory access (DMA) is a feature of computer systems that allows certain hardware subsystems to access main memory independently of the central processing unit (CPU). Without DMA, when the CPU is using programmed input/output, it is typically fully occupied for the entire duration of the read or write operation, and is thus unavailable to perform other work. With DMA, the CPU first initiates the transfer, then it does other operations while the transfer is in progress, and it finally receives an interrupt from the DMA controller when the operation is done. This feature is useful at any time that the CPU cannot keep up with the rate of data transfer, or when the CPU needs to perform work while waiting for a relatively slow I/O data transfer.  Many hardware systems use DMA, including disk drive controllers, graphics cards, network cards, sound cards.